Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
pass over section 3

Optician greatly simplifies the task of programming bijective
string lenses, but not all bidirectional transformations are
bijective.  For instance, \bibtex{} users are not typically interested
in preserving whitespace between words.  The order of author and title
fields is also likely irrelevant, and there may be equivalent ways of
writing the same name: ``Lovelace, Ada'' vs ``Ada Lovelace.''
Consequently, the following two \bibtex{} citations represent the same
logical object even though they differ in nonessential details.\saz{should we
  place such examples side-by-side?}
\begin{verbatim}
   @Book {Lovelace,
   Author = "Ada Lovelace",
   Title = {Notes},
   }

   @Book{     Lovelace,
   Title = {Notes},
   Author = "Lovelace, Ada",     }
\end{verbatim}

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
\ksf{Are these QRE combinators or QRE lens combinators?  
We are currently mushy about this distinction.  
The section needs a pass to clarify this point.
The headers of this section and the next imply that this section is
about QREs and the next section is about QRE lenses, but this section
is where we introduce the QRE lens combinators, right?  If so, we need
to do something to make the structure clearer.}\afm{I believe this are QRE
combinators, so we should be good.}

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{collapse} primitive to define this
whitespace-normalizing QRE.

\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}\bcp{Done, I think.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

One way to define the \codefont{c\_to\_s} function
is simply to write it from scratch in some ordinary programming language.
However, we can synthesize such functions automatically---here, \codefont{c\_to\_s} is the
get direction of a lens that can be synthesized using the 
\kw{synth} combinator:
%
\begin{lstlisting}
let $\ell$ = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = $\ell$.get 
\end{lstlisting}
\noindent \saz{I changed \codefont{get} $\ell$ to $\ell$.\kw{get} to be
  consistent with the earlier text.}
The first line above synthesizes a lens between \codefont{comma\_name} and
\codefont{space\_name} using the listed example transformation as a guide.  The
second line extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

@@

Figure~\ref{fig:example-qre} gives a QRE specification for the simple
\bibtex{} records we consider here.
\begin{figure}
\begin{lstlisting}
let wsp = [#\,\textvisiblespace\,#\n\t\r]+
let wsp_sp = collapse wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}\saz{This defines only one endpoint of the eventual
\bibtex{}--EndNote lens.  Maybe we should say ``definition'' instead of ``specification''?}
\label{fig:example-qre}
\end{figure}

>>>>

Optician greatly simplifies the task of programming bijective
string lenses, but not all bidirectional transformations are
bijective.  For instance, \bibtex{} users are not typically interested
in preserving whitespace between words.  The order of author and title
fields is also likely irrelevant, and there may be equivalent ways of
writing the same name: ``Lovelace, Ada'' vs ``Ada Lovelace.''
Consequently, the following two \bibtex{} citations represent the same
logical object even though they differ in nonessential details.

\begin{center}
  \begin{minipage}{2.2in}
    \centering
\begin{lstlisting}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Notes},
}
\end{lstlisting}
  \end{minipage}
  \begin{minipage}{2in}
    \centering
\begin{lstlisting}
@Book{     Lovelace,
Title = {Notes},
Author = "Lovelace, Ada",     }
\end{lstlisting}
  \end{minipage}
\end{center}

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{collapse} primitive to define this
whitespace-normalizing QRE.

\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
\end{lstlisting}
\iflastminute
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

One way to define the \codefont{c\_to\_s} function
is simply to write it from scratch in some ordinary programming language.
However, we can synthesize such functions automatically---here, \codefont{c\_to\_s} is the
get direction of a lens that can be synthesized using the 
\kw{synth} combinator:
%
\begin{lstlisting}
let $\ell$ = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = $\ell$.get 
\end{lstlisting}
\noindent
The first line above synthesizes a lens between \codefont{comma\_name} and
\codefont{space\_name} using the listed example transformation as a guide.  The
second line extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

@@

Figure~\ref{fig:example-qre} gives a QRE definition for the simple
\bibtex{} records we consider here.
\begin{figure}
\begin{lstlisting}
let wsp = [#\,\textvisiblespace\,#\n\t\r]+
let wsp_sp = collapse wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE definition of \bibtex{} records.}
\label{fig:example-qre}
\end{figure}
